Proof-Reconstruction in Propositional Logic

### 2017-10-19 paper/sections/abstract.tex
[ paper ] lots of grammar fixes.
[ paper ] working on corrections.

\begin{abstract}
In this paper, we describe a syntactical proof-reconstruction
approach to verify solutions generated by \Metis prover to problems in classical propositional logic without equality.
To verify these \Metis derivations, we translate them automatically
to \Agda proof-terms.
Using the proof-assistant, we were able to type-check the
formalization approach presented in this work. A formulation about
the inference rules used by \Metis.
This allowed us to type-check proof-objects step-by-step.

\keywords{proof-reconstruction, type theory, \Agda, automatic
theorem prover, classical propositional logic, \Metis.}

>>>>

\begin{abstract}
We describe a syntactical proof-reconstruction
approach to verify proofs generated by \Metis prover to problems in
classical propositional logic.
To verify these \Metis proofs, we developed a tool able to translate
such derivations to \Agda proof-terms.
We formulate in type theory each inference rule found
in \Metis derivations. This allowed us to type-check \Metis proofs
step-by-step. We leave it for future work,
the opportunity to extend this work to other provers, proof assistants, and other formalisms.

\keywords{proof-reconstruction, type theory,
automatic theorem prover}

### 2017-10-19 paper/sections/introduction.tex
[ paper ] working on corrections.

Proof-reconstruction is a hard labor since it depends on the
integration of two complex systems. On the one hand, we have the
automatic theorem provers (henceforth \ATP) and their specification
logic. These tools are usually classified in at least one of the
following categories.
A \SAT solver (e.g., \name{zChaff}~\cite{Moskewicz2001} and
\name{MiniSat}~\cite{Een2004}) to prove unsatisfiability of \CNF
formulas, a \abbre{QBF} solver (e.g.,
\name{GhostQ}~\cite{Klieber2014} and
\name{DepQBF}~\cite{Lonsing2017}) to prove satisfiability and
invalidity of quantified Boolean formulas, a \SMT solver
(e.g \name{CVC4}~\cite{Barrett2011}, \name{veriT}~\cite{bouton2009},
and \name{Z3}~\cite{DeMoura2008}) to prove unsatisfiability of
formulas from first-order logic with theories, and a prover for
validity of formulas from first-order logic with equality
(e.g., \name{E}~\cite{Schulz:AICOM-2002},
\name{leanCoP}~\cite{Otten2008},
\Metis~\cite{hurd2003first},
\name{SPASS}~\cite{Weidenbach2009} and
\name{Vampire}~\cite{Riazanov1999}), high-order logic (e.g.,
\name{Leo-II}~\cite{Benzmuller2008} and
\name{Satallax}~\cite{Brown2012}) or intuitionistic logic (e.g.,
\name{ileanCoP}~\cite{Otten2008},
\name{JProver}~\cite{Schmitt2001}, and
\name{Gandalf}~\cite{Tammet1997}), among others. On the other hand,
we have the proof checkers, interactive theorem provers (henceforth
\ITP) or proof assistants (e.g., \Agda~\cite{agdateam},
\name{Coq}~\cite{coqteam},
\name{Isabelle}~\cite{paulson1994isabelle}, and
\name{HOL4}~\cite{norrish2007hol}). The \ITP tools provide us the
logic framework to check and validate the reply of the \ATPs since
they allow us to define the formal language for the problems i.e.,
operators, logic variables, axioms, and theorems.

@@

In this paper, we ported a subset of the \Metis' inference rules to
\Agda, the propositional fragmented without equality. The formalization of
these rules allowed us to justify step-by-step \Metis' \TSTP derivations by
translating automatically to \Agda by a proof-reconstruction tool built for
this research named \Athena.
This paper has been organized in the following way.
The first section of the paper, Section~\ref{ssec:CPL}, will mention some
notation and syntax of the classical propositional logic.
In Section~\ref{ssec:metis-language-and-proofs}, we introduce the
\Metis prover and its deduction system.
In Section~\ref{sec:proof-reconstruction}, we present our
approach to reconstruct proofs deliver by \Metis in type theory.
Section~\ref{sec:conclusions} summarizes the main findings and includes a
discussion of future work.
Some related works to this paper are presented in Section~\ref{sec:related-work}.
Finally, in appendix \textsc{A}, we provide a small example of the
proof-reconstruction workflow.

>>>>

Proof-reconstruction is a non-trivial task since it depends on the
integration of two complex systems.

On the one hand, we have the
automatic theorem provers (henceforth \ATP) and their specification
logic. These tools are usually classified in at least one of the
following categories.
A \SAT solver (e.g., \name{zChaff}~\cite{Moskewicz2001} and
\name{MiniSat}~\cite{Een2004}) to prove unsatisfiability of \CNF
formulas, a \abbre{QBF} solver (e.g.,
\name{GhostQ}~\cite{Klieber2014} and
\name{DepQBF}~\cite{Lonsing2017}) to prove satisfiability and
invalidity of quantified Boolean formulas, a \SMT solver
(e.g \name{CVC4}~\cite{Barrett2011}, \name{veriT}~\cite{bouton2009},
and \name{Z3}~\cite{DeMoura2008}) to prove unsatisfiability of
formulas from first-order logic with theories, and a prover for
validity of formulas from first-order logic with equality
(e.g., \name{E}~\cite{Schulz:AICOM-2002},
\name{leanCoP}~\cite{Otten2008},
\Metis~\cite{hurd2003first},
\name{SPASS}~\cite{Weidenbach2009} and
\name{Vampire}~\cite{Riazanov1999}), high-order logic (e.g.,
\name{Leo-II}~\cite{Benzmuller2008} and
\name{Satallax}~\cite{Brown2012}) or intuitionistic logic (e.g.,
\name{ileanCoP}~\cite{Otten2008},
\name{JProver}~\cite{Schmitt2001}, and
\name{Gandalf}~\cite{Tammet1997}), among others. On the other hand,
we have the proof checkers, interactive theorem provers (henceforth
\ITP) or proof assistants (e.g., \Agda~\cite{agdateam},
\name{Coq}~\cite{coqteam},
\name{Isabelle}~\cite{paulson1994isabelle}, and
\name{HOL4}~\cite{norrish2007hol}). The \ITP tools provide us the
logic framework to check and validate the reply of the \ATPs since
they allow us to define the formal language for the problems i.e.,
operators, logic variables, axioms, and theorems.

@@

In this paper, we ported a subset of the \Metis' inference rules to
\Agda, the propositional fragmented. The formalization of
these rules allowed us to justify step-by-step \Metis' \TSTP derivations by
translating automatically to \Agda by a proof-reconstruction tool built for
this research named \Athena.
This paper has been organized in the following way.
In Section~\ref{ssec:CPL}, we introduce some
notation and syntax of propositional logic.
In Section~\ref{ssec:metis-language-and-proofs}, we introduce the
\Metis prover and its deduction system.
In Section~\ref{sec:proof-reconstruction}, we present our
approach to reconstruct proofs deliver by \Metis in type theory.
Section~\ref{sec:conclusions} summarizes the main findings and includes a
discussion of future work.
Some related works to this paper are presented in Section~\ref{sec:related-work}.
Finally, in appendix \textsc{A}, we provide a small example of the
proof-reconstruction workflow.

### 2017-10-19 paper/sections/metis-language-and-proofs.tex
[ paper ] working on corrections.

\Metis is an automatic theorem prover written in \len{Standard ML}
for first-order logic with equality developed by John Hurd
\cite{hurd2003first}. This prover is suitable for proof
reconstruction since it provides well-documented proof-objects to
justify its deduction steps from the basis of only six inference
rules. There are already some tools that perform proof-reconstruction
with \Metis. %Some of these systems are the following.

@@

The \name{rule} stands for the inference name; the other fields are
supporting arguments or useful information to apply the reasoning
step, and the list of parents nodes. For instance, in Fig.
\ref{fig:metis-proof-tstp}, \strip is the name of one of the
inference with no arguments and only one parent node, the
\name{goal}.

\begin{figure}
\begin{verbatim}
fof(premise, axiom, p).
fof(goal, conjecture, p).
fof(subgoal_0, plain, p, inference(strip, [], [goal])).
fof(negate_0_0, plain, ~ p, inference(negate, [], [subgoal_0])).
fof(normalize_0_0, plain, ∼ p,
  inference(canonicalize, [], [negate_0_0])).
fof(normalize_0_1, plain, p,
  inference(canonicalize, [], [premise])).
fof(normalize_0_2, plain, $false,
  inference(simplify, [], [normalize_0_0, normalize_0_1]))
cnf(refute_0_0, plain, $false,
  inference(canonicalize, [], [normalize_0_2])).
\end{verbatim}
\caption{The \texttt{Metis}' \texttt{TSTP} derivation for the
problem $p\vdash p$.}
\label{fig:metis-proof-tstp}
\end{figure}

@@

\begin{figure}
\centering
  \begin{bprooftree}\tt
    \AxiomC{}
    \RightLabel{negate}
    \UnaryInfC{$\neg p$}
    \RightLabel{strip}
    \UnaryInfC{$\neg p$}
    \AxiomC{}
    \RightLabel{axiom}
    \UnaryInfC{$p$}
    \RightLabel{canonicalize}
    \UnaryInfC{$p$}
    \RightLabel{simplify}
    \BinaryInfC{$⊥$}
    \RightLabel{canonicalize}
    \UnaryInfC{$⊥$}
  \end{bprooftree}
  \caption{The \Metis' proof tree for $p \vdash p$ from the
  derivation in Fig.~\ref{fig:metis-proof-tstp}}
  \label{fig:metis-example}
\end{figure}

A proof-object delivered by \Metis encodes a natural
deduction proof. The deduction system uses six inference rules
\cite{hurd2003first} listed in Fig.~\ref{fig:metis-inferences}.
With these inference rules as the only valid deduction steps, \Metis
attempts to prove conjectures by refutation (see. e.g.,
\emph{falsium} in the last conclusion of the \TSTP derivation in
Fig.~\ref{fig:metis-example}).

>>>>

\Metis is an automatic theorem prover written in \len{ML}
for first-order logic with equality developed by John Hurd
\cite{hurd2003first}. This prover is suitable for proof
reconstruction since it provides well-documented proof-objects to
justify its deduction steps from the basis of only six inference
rules. There are already some tools that perform proof-reconstruction
with \Metis. %Some of these systems are the following.

@@

The \name{rule} stands for the inference name; the other fields are
supporting arguments or useful information to apply the reasoning
step, and the list of parents nodes. For instance, in Fig.
\ref{fig:metis-proof-tstp}, \strip for example is the name of one of the inference with no arguments and only one parent node, the
\name{goal}.

\begin{figure}
\begin{verbatim}
fof(premise, axiom, p).
fof(goal, conjecture, p).
fof(subgoal_0, plain, p, inference(strip, [], [goal])).
fof(negate_0_0, plain, ~ p, inference(negate, [], [subgoal_0])).
fof(normalize_0_0, plain, ∼ p,
  inference(canonicalize, [], [negate_0_0])).
fof(normalize_0_1, plain, p,
  inference(canonicalize, [], [premise])).
fof(normalize_0_2, plain, $false,
  inference(simplify, [], [normalize_0_0, normalize_0_1]))
cnf(refute_0_0, plain, $false,
  inference(canonicalize, [], [normalize_0_2])).
\end{verbatim}
\caption{\texttt{Metis}' \texttt{TSTP} derivation for the
problem $p\vdash p$.}
\label{fig:metis-proof-tstp}
\end{figure}

@@

\begin{figure}
\centering
  \begin{bprooftree}\tt
    \AxiomC{}
    \RightLabel{negate}
    \UnaryInfC{$\neg p$}
    \RightLabel{strip}
    \UnaryInfC{$\neg p$}
    \AxiomC{}
    \RightLabel{axiom}
    \UnaryInfC{$p$}
    \RightLabel{canonicalize}
    \UnaryInfC{$p$}
    \RightLabel{simplify}
    \BinaryInfC{$⊥$}
    \RightLabel{canonicalize}
    \UnaryInfC{$⊥$}
  \end{bprooftree}
  \caption{\Metis' proof tree for $p \vdash p$. Derivation in Fig.~\ref{fig:metis-proof-tstp}}
  \label{fig:metis-example}
\end{figure}

The output of a proof delivered by \Metis encodes a natural
deduction proof. The deduction system uses six inference rules
\cite{hurd2003first} listed in Fig.~\ref{fig:metis-inferences}.
With these inference rules as the only valid deduction steps, \Metis
attempts to prove conjectures by refutation (e.g.,
\emph{falsium} in the last conclusion of the \TSTP derivation in
Fig.~\ref{fig:metis-example}).

### 2017-10-19 paper/sections/preliminaries.tex
[ paper ] working on corrections.

\textit{Propositions.}
We define the classical propositional logic as usual.
A proposition is an expression of indivisible propositional atoms
(e.g., symbols $φ₀, φ₁, \cdots$),
binary logic connectives ($∧, ∨, ⇒, ⇔)$,
propositional negation $(¬)$, and logic constants $⊥$
and $⊤$. We use the inductive definition of the set \Prop
presented in \cite{VanDalen1994} as the mechanism to produce any
propositional formula.
\begin{itemize}
\item $⊥, ⊤ \in \Prop$.
\item $ϕ_i \in \Prop$ for all $i \in \mathbb{N}$.
\item If $φ, ψ\in \Prop$, then $(φ\ \square\ ψ) \in \Prop$,
where the $\square$ symbol is one of the binary connectives mentioned above.
\item If $φ \in \Prop$, then $(¬ φ) \in \Prop$.
\end{itemize}

\emph{Derivations.}
A derivation of $φ$ is a mechanism to obtain the conclusion $φ$ from
the application of some deduction steps. Such a deduction step has
the form of a set of propositional \emph{premises} that produces
one or more proposition as the conclusion.

As we show in the following
figure, the premises are always above the line, the conclusions
are below it, and the name of the inference or the theorem used is
on the left or on the right of the line. As a notation convention, if the name of a rule used in the derivation step is typed using verbatim font face, it means the inference rule belongs to our implementation~\cite{AgdaProp, AgdaMetis} in the Proof Assistant of a inference rule presented
with a similar name.

\begin{equation*}
\label{eq:derivation}
\begin{bprooftree}
  \AxiomC{$φ$}
  \AxiomC{$ψ$}
  \RightLabel{∧-intro.}
  \BinaryInfC{$φ ∧ ψ$}
\end{bprooftree}
% \qquad
% \begin{bprooftree}
%   \AxiomC{$[¬ φ]$}
%   \noLine
%   \UnaryInfC{$\vdots$}
%   \noLine
%   \UnaryInfC{$⊥$}
%   \RightLabel{\abbre{RAA}}
%   \UnaryInfC{φ}
% \end{bprooftree}
\end{equation*}

\vskip 2mm

@@

\caption{Inference rules in classical propositional logic.}
\label{fig:CPL-inference-rules}
\end{figure}

>>>>

\textit{Propositions.}
We define the propositional logic as usual.
A proposition is an expression of indivisible propositional atoms
(e.g., symbols $φ₀, φ₁, \cdots$), the logic constants:
$⊥$, $⊤$, the binary connectives ($∧, ∨, ⇒, ⇔)$, and the negation
$(¬)$. We use the inductive definition of the set \Prop
presented in \cite{VanDalen1994} as the mechanism to produce any
propositional formula.

\begin{itemize}
\item $⊥, ⊤ \in \Prop$.
\item $ϕ_i \in \Prop$ for all $i \in \mathbb{N}$.
\item If $φ, ψ\in \Prop$, then $(φ\ \square\ ψ) \in \Prop$,
where the $\square$ symbol is one of the binary connectives mentioned above.
\item If $φ \in \Prop$, then $(¬ φ) \in \Prop$.
\end{itemize}

\emph{Derivations.}
A derivation of $φ$ is a mechanism to obtain the conclusion $φ$ from
the application of some deduction steps. Such deduction step has
the form of a set of propositional \emph{premises} that produces
propositions as the conclusion.

As we show in Fig.~\ref{fig:derivation}, the premises are always above the line, the conclusion
is below it, and the name of the inference or the theorem used is
on the left or on the right of the line.
% As a notation convention, if the name of a rule used in the derivation step is typed using verbatim font face, for example, \verb!resolve!, it means the inference rule belongs to our implementation~\cite{AgdaProp, AgdaMetis} in the Proof Assistant of a inference rule presented
% with a similar name.

\begin{figure}
\label{fig:derivation}
\begin{equation*}
\begin{bprooftree}
  \AxiomC{$φ$}
  \AxiomC{$ψ$}
  \RightLabel{∧-intro}
  \BinaryInfC{$φ ∧ ψ$}
\end{bprooftree}
\end{equation*}
\caption{Derivation example.}
\end{figure}

@@

\[
\begin{bprooftree}
\AxiomC{}
\RightLabel{PEM}
\UnaryInfC{$Γ ⊢ φ ∨ ¬ φ$}
\end{bprooftree}\qquad
\begin{bprooftree}
\AxiomC{$Γ, ¬ φ ⊢ ⊥$}
\RightLabel{RAA}
\UnaryInfC{$Γ ⊢ φ$}
\end{bprooftree}
\]

\caption{Inference rules for propositional logic.}
\label{fig:CPL-inference-rules}
\end{figure}

### 2017-10-19 paper/sections/proof-reconstruction.tex
[ paper ] working on corrections.

The proof-reconstruction approach proposed here consists of a series
of steps similar to the workflow presented by \citeauthor{sultana2015} in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is a proof
assistant, \Agda. We choose \Agda, but another proof assistant with the same
support of type theory and inductive types is valid.
%  (see a description of \Agda for its main features in
% Section~\ref{sec:implementation}).
The overview of the proof-reconstruction is presented in
Fig.~\ref{fig:proof-reconstruction-workflow}, the process begin with
a \TPTP file that encodes a problem in \CPL. We use this file as the
input of the \Metis prover and if the problem is a theorem, \Metis can reply
with a derivation in \TSTP format of the proof.

@@

In the remainder part of this section, a formal description using
type theory is presented to build definitions and theorems of functions
necessary to emulate \Metis' inference rules. As a convention,
We write the type judgments as ``$a : A$'' to denote that $a$
has $\abbre{A}$ type. Types are written using small caps.
We use \Prop type for propositions and the dependent type, $Γ ⊢ φ$ for
theorems where $Γ$ has $\abbre{List}~\Prop$ type, and $φ$ has type $\Prop$.
$\abbre{List}$ type is the usual inductive type for lists. For syntactical
equivalence, we write $φ ≡ ψ$ for $φ : \Prop$ and $ψ : \Prop$.

@@

Henceforth, a natural number has \Nat type, and it is defined as
usual, i.e., \zero and \suc are the data constructors and some
arithmetic operations are defined: ($+,\ -,\ *$). We include syntax
sugar for $\zero$, $\suc$, $\suc(\zero)$, $\cdots$, with the
classical representation: $1$, $2$, $\cdots$ as well.

>>>>

The proof-reconstruction approach proposed consists of a series
of steps similar to the workflow presented by \citeauthor{sultana2015} in
\cite{sultana2015}. This process is a translation from a source
system to a target system. In our case, the system of origin, the
automatic theorem prover, is \Metis; the target system is a proof
assistant, \Agda. We choose \Agda, but another proof assistant with the same support of type theory and inductive types can be used.
%  (see a description of \Agda for its main features in
% Section~\ref{sec:implementation}).
The overview of the proof-reconstruction is presented in
Fig.~\ref{fig:proof-reconstruction-workflow}. The process begin with
a \TPTP file that encodes a problem in \CPL. We use this file as the
input of the \Metis prover and if the problem is a theorem, \Metis can reply a derivation in \TSTP format of the proof.

@@

In the remainder part of this section, a formal description using
type theory is presented to build definitions and theorems of
functions necessary to emulate \Metis' inference rules.
As a convention, we write the type judgments as ``$m : M$'' to denote that $m$ has $\abbre{M}$ type.
Types are written using small caps.
We use \Prop type for propositions and the dependent type, $Γ ⊢ φ$ for
theorems where $Γ$ has $\abbre{List}~\Prop$ type, and $φ$ has type $\Prop$.
$\abbre{List}$ type is the usual inductive type for lists. For syntactical
equivalence, we write $φ ≡ ψ$ for $φ : \Prop$ and $ψ : \Prop$.

@@

Henceforth, a natural number has \Nat type, and it is defined as
usual, i.e., \zero and \suc are the data constructors and some
arithmetic operations are defined: ($+,\ -,\ *$). We include syntax
sugar for $\zero$, $\suc$, $\suc(\zero)$, $\cdots$, with the
classical representation: $0$, $1$, $2$, $\cdots$ as well.
